Evaluating the Performance and Energy Efficiency of COSMO-ART, a Fully Online Coupled Model System with Numerical Weather Forecast and Chemical Transport Models

### 2014-04-07 paper_enahpc14/Sec2:Power-measurement.tex
adds Pilatus results

\subsection{Hardware description}
\label{subsec:2.1}

\subsubsection{Monch (CSCS - ETH Zurich)}
While hardware  platforms mature  and are replaced,  we have  chosen a
state-of-the-art  dual-socket Intel  Ivybridge-based  platform (called
``Monch'')  for our power  measurements, as  it is  slated to  stay in
service without upgrade for the duration of the Exa2Green project.  In
principle at least,  this architecture could be recreated  or found in
an identical configuration beyond  the lifetime of the project.  Given
that  the baseline  benchmark  can be  reproduced  within an  expected
variance (see subsection~\ref{subsec:1.2}),  and that the baseline run
configuration can be  used in all future versions of  the code, a fair
comparison  will  be  made  between  the baseline  and  the  milestone
versions of COSMO-ART.

Monch, which is installed  at the Swiss National Supercomputing Center
(CSCS)  of  the  ETH  Zurich,  is a  10  rack  NEC-provided  (LX-2400)
Intel-based cluster, utilised  from people that are part  of the Swiss
Platform  for  Advanced Scientific  Computing  (PASC),  which aims  at
positioning  Swiss  computational sciences  in  the emerging  exascale
era. It is  composed of 16 administration nodes,  312 standard compute
nodes, 24 large-memory compute nodes, 24 huge-memory compute nodes and
8  servers to  provide  I/O  to Lustre.   Each  standard compute  node
comprises 2 Intel 10-core Ivy Bridge EP E5-2660v2 processors operating
at  2.2 GHz,  offering 32GB  of DDR3  1600MHz RAM,  with a  base clock
frequency of 2.2 GHz, themselves  connected by a high speed Infiniband
network,  based on Mellanox  SX6036 managed  FDR (Fourteen  Data Rate)
switches, with a 56 Gb/s speed.

@@

\subsubsection{University of Hamburg}
To assess the  performance and the energy efficiency  of COSMO-ART, we
employ   a    version   of   the    integrated   framework   presented
in~\cite{energy13}  that  works in  combination  with VampirTrace  and
Vampir,   which  are   profiling/tracing   and  visualization   tools,
respectively.
%The left part of the \vref{fig:Lustre} offers a graphical representation of
%the Lustre architecture; the right depicts the tracing and profiling framework.
To  use our  approach,  COSMO-ART is  compiled  using the  VampirTrace
compiler wrappers, which automatically  instrument the Fortran code of
the  model. Next,  COSMO-ART is  run  on the  nodes, thus  dissipating
certain  amount of  power.  The  server nodes  are connected  to power
measurement  devices that  account for  the  dissipated power/consumed
energy and  send the power data  to the tracing  server.  The attached
VampirTrace \pmlib plugin employs the client API that sends start/stop
primitives in order to gather captured data by the wattmeters onto the
tracing server,  where an  instance of the  \pmlib server  is running.
Once COSMO-ART run is finished, the VampirTrace \pmlib plugin receives
the  power   data  from  the  tracing   server.   The  instrumentation
post-process  generates the  performance  trace files  and the  \pmlib
plugin inserts the power data into them.

>>>>

\subsection{Hardware description}
\label{subsec:2.1}
While hardware  platforms mature  and are replaced,  we have  chosen a
state-of-the-art  Intel's third-generation  Core (aka  ``Ivy Bridge'')
processing platform (called ``Monch'')  for our power measurements, as
it  is slated  to stay  in service  without hardware  upgrade  for the
duration  of  the Exa2Green  project.   In  principle  at least,  this
architecture could be recreated or found in an identical configuration
beyond the lifetime of the project.  Given that the baseline benchmark
can    be    reproduced    within    an   expected    variance    (see
subsection~\ref{subsec:1.2}), and that  the baseline run configuration
can be used in all future versions of the code, a fair comparison will
be made between the baseline  and the milestone versions of COSMO-ART.
In  this   study,  a  complementary   energy-to-solution  benchmarking
comparison is carried out on  the ``Pilatus'' cluster based on Intel's
previous-generation  ``Sandy  Bridge'' processors,  known  to be  more
power consuming.

\subsubsection{Monch (CSCS - ETH Zurich)}
Monch, which is installed  at the Swiss National Supercomputing Center
(CSCS)  of the ETH  Zurich, is  a 10  rack NEC-provided  (LX-2400) and
dual-socket Intel-based cluster, utilised from people that are part of
the   Swiss  Platform   for  Advanced   Scientific   Computing  (PASC,
\url{http://www.pasc-ch.org/}).   It  is   composed  of  312  standard
compute  nodes,  24  large-memory  compute nodes  and  24  huge-memory
compute nodes.   Each standard compute node comprises  2 Intel 10-core
Ivy Bridge EP E5-2660v2 processors operating at 2.2 GHz, offering 32GB
of DDR3 1600MHz  RAM, themselves connected by a  high speed Infiniband
network,  based on Mellanox  SX6036 managed  FDR (Fourteen  Data Rate)
switches, with a 56 Gb/s speed.

\subsubsection{Pilatus (CSCS - ETH Zurich)}
Pilatus is a dual-socket Intel  Sandy-Bridge based cluster used as Piz
Daint pre-post processing cluster.  It is composed of 42 compute nodes
with  a total  number of  1344 cores  in hyper-threading  and  2688 GB
aggregate memory and has 2  high-speed interconnects based on FDR: the
first is  dedicated to the MPI  traffic and the second  to the storage
high  speed traffic.  The  2 login  nodes  and the  42 computes  nodes
consists in  11 twin-pair Intel  E5-Series DALCO r2264i4t  2U scalable
compute  modules.  Each  module contains  4 compute  nodes based  on 2
Intel  8-core Intel  Xeon  E5-2670 processors  operating  at 2.6  GHz,
offering  64GB of  DDR3 1600MHz  RAM, themselves  connected by  a high
speed  Infiniband  network,  based  on  Mellanox  SX6036  managed  FDR
(Fourteen Data Rate) switches, with a 56 Gb/s speed.

@@

\subsubsection{University of Hamburg}
To assess the  performance and the energy efficiency  of COSMO-ART, we
employ   a    version   of   the    integrated   framework   presented
in~\cite{energy13}  that  works in  combination  with VampirTrace  and
Vampir,   which  are   profiling/tracing   and  visualization   tools,
respectively.
%The left part of the \vref{fig:Lustre} offers a graphical representation of
%the Lustre architecture; the right depicts the tracing and profiling framework.
To  use our  approach,  COSMO-ART is  compiled  using the  VampirTrace
compiler wrappers, which automatically  instrument the Fortran code of
the  model. Next,  COSMO-ART is  run  on the  nodes, thus  dissipating
certain  amount of  power.  The  server nodes  are connected  to power
measurement  devices that  account for  the  dissipated power/consumed
energy and  send the power data  to the tracing  server.  The attached
VampirTrace \pmlib plugin employs the client API that sends start/stop
primitives in order to gather captured data by the wattmeters onto the
tracing server,  where an  instance of the  \pmlib server  is running.
Once COSMO-ART run is finished, the VampirTrace \pmlib plugin receives
the  power   data  from  the  tracing   server.   The  instrumentation
post-processing generates  the performance trace files  and the \pmlib
plugin inserts the power data into them.

### 2014-04-07 paper_enahpc14/Sec3:Evaluation.tex
adds Pilatus results

\subsection{Sofware environment}
\label{subsec:3.1}
The  software environment  on Monch  is controlled  using  the modules
framework which gives an easy  and flexible mechanism to access to all
of  the  CSCS  provided  compilers,  tools and  applications.   It  is
particularly  useful for  testing code  portability  between different
compilers  or for  changing  between different  versions  of the  same
compiler.  INT2LM  and the COSMO-ART model are  implemented in Fortran
90 for distributed memory parallel computers using the Message Passing
Interface (MPI).  For  our initial benchmarking, we opted  for the GNU
compiler (gcc/4.8.1) with the -O3  compiler flag as it generally gives
a  good  level  of optimization  and  the  code  runs faster  in  this
configuration  than  when  compiled  with  the  intel  compiler,  also
available on  Monch.  Besides, we installed  the MPICH2 implementation
of MPI (mvapich2/1.9) as well  as the commonly used HDF5 (hdf5/1.8.12)
and NetCDF  (netcdf/4.3.1) libraries in favor of  the traditional GRIB
library  for  the  management  of  extremely large  and  complex  data
collections.    Note   that   the  Linux   2.6.32-358.11.1.el6.x86\_64
operating system was used for all the Monch compute nodes.

@@

A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was  produced and run on a  full rack of 52  nodes on Monch
(monchc[029-080]),  which represents a  total of  1040 cores  using 20
tasks per node.  The calculated region was mapped to the participating
processors  using a 2D-partitioning  strategy. The  distribution along
the  x and  y coordinates  is defined  in the  namelist  INPUT\_ORG by
setting:  $nprocx=40$   and  $nprocy=26$,  as  the   total  number  of
processors has to be equal to $nprocx \times nprocy$.  Besides we take
$nprocio=0$.   Hyperthreading  is  not  considered in  this  study  as
previous attempts  of its  use revealed that  it always led  to higher
energy-to-solution.\\

@@

\subsection{Power-measurement results}
\label{subsec:3.3}
Measuring  the power  consumption on  one entire  Cray cabinet  is the
easiest way to calculate energy-to-solution for given applications.

\begin{figure*}
  \includegraphics[width=0.4\textwidth]{Figs/NRJ_benchmark_Monch.eps}
  \caption{Isola E1 Rack 2 Total Power and Isola E1 Total Power}
  \label{fig:1}
\end{figure*}

The run  was issued successfully  twice, the start time  and execution
time for both jobs are:
\begin{itemize}
\item start=12:51:03, end=13:20:46 $\Rightarrow$ T = 00:29:43 = 1783 s
\item start=14:28:40, end=14:58:02 $\Rightarrow$ T = 00:29:22 = 1762 s
\end{itemize}

@@

The  total  energy to  solution  is  the straightforward  calculation:\\

Energy-to-solution  =  Integral  of  power consumption  over  the  job
duration

Often  the  average  power  consumption  times  the  job  duration  is
sufficient.

@@

>>>>

\subsection{Sofware environment}
\label{subsec:3.1}
The software environment on Monch  and Pilatus is controlled using the
modules framework which gives an easy and flexible mechanism to access
to all of the CSCS  provided compilers, tools and applications.  It is
particularly  useful for  testing code  portability  between different
compilers  or for  changing  between different  versions  of the  same
compiler.  INT2LM  and the COSMO-ART model are  implemented in Fortran
90 for distributed memory parallel computers using the Message Passing
Interface (MPI).  For  our initial benchmarking, we opted  for the GNU
compiler  (gcc/4.8.1 on  Monch,  gcc/4.8.2 on  Pilatus)  with the  -O3
compiler flag as  it generally gives a good  level of optimization and
the code runs faster in this configuration than when compiled with the
intel compiler  (intel/14.0.1), also available on  Monch.  Besides, we
installed the  MPICH2 implementation of MPI (mvapich2/1.9)  as well as
the  commonly  used   HDF5  (hdf5/1.8.12)  and  NetCDF  (netcdf/4.3.1)
libraries in favor of the  traditional GRIB library for the management
of  extremely  large and  complex  data  collections.   Note that  all
compute  nodes of  Monch  and Pilatus  used  respectively the  ``Linux
2.6.32-358.11.1.el6.x86\_64''   and   ``Linux   3.0.101-0.15-default''
operating systems.

@@

A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was  produced and run on a  full rack of 52  nodes on Monch
(monchc[029-080]), i.e.  a total of 1040 cores using 20 tasks per node
and on  a full rack of  42 nodes on Pilatus  (pilatus[03-44]), i.e.  a
total of  1344 cores using 16  tasks per node.   The calculated region
was  mapped to  the participating  processors using  a 2D-partitioning
strategy.  The distribution along the  x and y coordinates was defined
by setting: $nprocx=40$ and  $nprocy=26$ for Monch and $nprocx=28$ and
$nprocy=24$.  Besides as this version of COSMO-ART doesn't make use of
the   GRIB   library,  we   specified   $nprocio=0$   for  GRIB   I/O.
Hyperthreading is not considered in this study as previous attempts of
its use revealed that it always led to higher energy-to-solution.\\

@@

\subsection{Power-measurement results}
\label{subsec:3.3}
Measuring  the power  consumption on  one entire  Cray cabinet  is the
easiest    way    to    calculate   energy-to-solution    for    given
applications.  The total  energy  to solution  is the  straightforward
calculation:\\

Energy-to-solution  =  Integral  of  power consumption  over  the  job
duration\\

However, the average power consumption times the job duration is often
sufficient.

\subsubsection{Monch (CSCS - ETH Zurich)}
\begin{figure*}
  \includegraphics[width=0.4\textwidth]{Figs/NRJ_benchmark_Monch.eps}
  \caption{Isola E1 Rack 2 Total Power and Isola E1 Total Power}
  \label{fig:1}
\end{figure*}

The 1-day simulation was issued successfully twice, the start time and
execution time for both runs are:
\begin{itemize}
\item start=12:51:03, end=13:20:46 $\Rightarrow$ T = 00:29:43 = 1783 s
\item start=14:28:40, end=14:58:02 $\Rightarrow$ T = 00:29:22 = 1762 s
\end{itemize}

@@

@@

\subsubsection{Pilatus (CSCS - ETH Zurich)}
\begin{figure*}
  \includegraphics[width=0.4\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
  \caption{Isola HD Pilatus Total Power}
  \label{fig:2}
\end{figure*}

The 1-day simulation was issued successfully four times and the 2-days
run only once, the start time and execution time for all jobs are:
\begin{itemize}
\item start=13:07:25, end=13:29:47 $\Rightarrow$ T = 00:22:22 = 1342 s
\item start=13:46:04, end=14:08:30 $\Rightarrow$ T = 00:22:26 = 1346 s
\item start=14:27:51, end=14:50:01 $\Rightarrow$ T = 00:22:10 = 1330 s
\item start=15:10:04, end=15:32:17 $\Rightarrow$ T = 00:22:13 = 1333 s
\item start=15:41:48, end=16:26:13 $\Rightarrow$ T = 00:44:25 = 2665 s
\end{itemize}

Again, the power  consumption for all jobs is  calculated by averaging
the  corresponding  power  measurements  from the  Excel  file.  Power
results  account  for  the  Isola  HD Pilatus  Total  Power.  As  time
resolution  is 5  minutes for  the output  results, the  average power
consumption is computed by considering  4 values for each single 1-day
run and 9 values for the 2-days run (see Table~\ref{tab:2}).

\begin{table}
  \begin{center}
    \caption{}
    \label{tab:2}
    \begin{tabular}{lll}
      \hline\noalign{\smallskip}
      Time & Measured power (W) & Average power (W)  \\
      \noalign{\smallskip}\hline\noalign{\smallskip}
       13:10:00 & 1.7824080000e+04 & 1.812201417e+04 \\ 
       13:15:00 & 1.8293650000e+04 &  \\ 
       13:20:00 & 1.8125140000e+04 &  \\ 
       13:25:00 & 1.8245186667e+04 &  \\ 
      \noalign{\smallskip}\hline\noalign{\smallskip}
       13:50:00 & 1.7849206667e+04 & 1.797961083e+04 \\ 
       13:55:00 & 1.7977790000e+04 &  \\ 
       14:00:00 & 1.8030646667e+04 &  \\ 
       14:05:00 & 1.8060800000e+04 &  \\ 
      \noalign{\smallskip}\hline\noalign{\smallskip}
       14:30:00 & 1.7868826667e+04 & 1.806545167e+04 \\ 
       14:35:00 & 1.8115786667e+04 &  \\ 
       14:40:00 & 1.8188520000e+04 &  \\ 
       14:45:00 & 1.8088673333e+04 &  \\ 
      \noalign{\smallskip}\hline\noalign{\smallskip}
       15:15:00 & 1.7857800000e+04 & 1.797302833e+04 \\ 
       15:20:00 & 1.7962513333e+04 &  \\ 
       15:25:00 & 1.8078226667e+04 &  \\ 
       15:30:00 & 1.7993573333e+04 &  \\ 
      \noalign{\smallskip}\hline\noalign{\smallskip}
       15:45:00 & 1.7801980000e+04 & 1.799757815e+04 \\ 
       15:50:00 & 1.7925386667e+04 &  \\ 
       15:55:00 & 1.8076986667e+04 &  \\ 
       16:00:00 & 1.8082966667e+04 &  \\ 
       16:05:00 & 1.7903206667e+04 &  \\ 
       16:10:00 & 1.8063913333e+04 &  \\ 
       16:15:00 & 1.8355030000e+04 &  \\ 
       16:20:00 & 1.7971580000e+04 &  \\ 
       16:25:00 & 1.7797153333e+04 &  \\ 
      \noalign{\smallskip}\hline
    \end{tabular}
  \end{center}
\end{table}

Thus, the total energy consumption corresponds to: 
\begin{itemize}
\item E = 1342 x 18122.01417 = 24319743.01614 J $\sim$ 24.32 MJ per day of simulation
\item E = 1346 x 17979.61083 = 24200556.17718 J $\sim$ 24.20 MJ per day of simulation
\item E = 1330 x 18065.45167 = 24027050.72110 J $\sim$ 24.03 MJ per day of simulation
\item E = 1333 x 17973.02833 = 23958046.76389 J $\sim$ 23.96 MJ per day of simulation
\item E = 2665 x 17997.57815 = 47963545.76975 J $\sim$ 47.96 MJ per day of simulation
\end{itemize}
